---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# cutr

cutr extends `base::cut.default`'s possibilities, getting inspiration from existing 
alternatives such as `Hmisc::cut2` and the `ggplot2::cut_*` family of functions, 
but going much further.

## Installation

You can install cutr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("moodymudskipper/cutr")
```

## default behavior + `closed` and `open_end`

We build a distribution and show the default behavior, in the rest of the
document we'll show results using `table` as it's more telling.

```{r}
library(cutr)
x <- c(rep(1,3),rep(2,2),3:6,17:20)
hist(x,breaks = 0:20)
cuts <- c(0,5,10,20)
smart_cut(x,cuts)
table(smart_cut(x,cuts))
```

The output is very similar to what we get with `base` or `Hmisc` and `ggplot2` 
alternatives.

```{r}
table(base::cut(x,cuts))
table(Hmisc::cut2(x,cuts))
table(smart_cut(x,cuts,"breaks"))
```

A the difference with `base` is that we close by default the left size,  this can be changed by setting the `closed` parameter to `"right"`.

Another is that the ends are both closed in `smart_cut`, this can be changed by setting the `open_end` parameter to `FALSE`

`closed` is borrowed from `ggplot2::cut_*` functions, and corresponds to the `RIGHT` parameter of `base::cut.default`.


`open_end` corresponds to the (negated) `include.lowest` parameter of `base::cut`.

The difference with `Hmisc` is due to the formatting function, which is `formatC` for `cut` and `smart_cut`, and `format` for `Hmisc` (which gives all labels the same width). The display of `smart_cut` is highly flexible thanks to the argument `format_fun` detailed further in this document.

## `what` and `i`

The `what` parameter determines how cuts will be chosen, depending on the value of `i`.

```{r}
table(smart_cut(x,cuts,"breaks"))                   # fixed breaks
table(smart_cut(x,2,"groups"))                      # groups defined by quantiles
table(smart_cut(x,list(2,"balanced"),"groups"))     # optimized groups of equal size
table(smart_cut(x,3,"n_by_group"))                  # try to get 3 items by group using quantiles
table(smart_cut(x,list(3,"balanced"),"n_by_group")) # try to get 3 items by group using optimization
table(smart_cut(x,3,"n_intervals"))                 # intervals of equal width
table(smart_cut(x,7,"width"))                       # interval of equal defined width, start on 1st value
table(smart_cut(x,list(7,"right"),"width"))         # interval of equal defined width, end on last value
table(smart_cut(x,list(6,"centered"),"width"))      # interval of equal defined width, centered
table(smart_cut(x,list(6,"centered0"),"width"))     # interval of equal defined width, centered on 0
table(smart_cut(x,list(7,0),"width"))               # interval of equal defined width, starting on 0
table(smart_cut(x,3,"cluster"))                     # create groups by running a kmeans clustering
```



## `simplify`

`TRUE` by default, when a value is the only one in its group, display it as a 
label, without brackets.  Similar to `oneval` in `Hmisc::cut2`.

```{r}
table(smart_cut(x, 5, "width", simplify = TRUE))
table(smart_cut(x, 5, "groups", simplify = FALSE))
```


## `expand`

expand makes sure all values from x will be in an interval by expanding the cut
points. `base::cut.default` never expands, `Hmisc::cut2` always expands.

```{r}
table(smart_cut(x,c(4,10,18)))
table(smart_cut(x,c(4,10,18),expand = FALSE))
```

## `crop`

`crop` is `FALSE` by default, if `TRUE` the side intervals are reduced to fit the data.

```{r}
table(smart_cut(x,c(0,10,30)))
table(smart_cut(x,c(0,10,30),crop = TRUE))
```

## `squeeze`

`squeeze` is `FALSE` by default, if `TRUE` every interval is reduced to fit the data.

```{r}
table(smart_cut(x,c(0,10,30)))
table(smart_cut(x,c(0,10,30),squeeze = TRUE))
```


## `brackets` + `sep`

Different brackets can be chosen

```{r}
table(smart_cut(x,c(0,10,30), brackets = c("]","[","[","]")))
table(smart_cut(x,c(0,10,30), brackets = NULL, sep = "~", squeeze= TRUE))
```

## `labels`

`labels` can be a vector just like in `base::cut.default`, but it can also
be a function of 2 arguments, which are a vector of values contained in the 
interval and a vector of cutpoints.


```{r}
table(smart_cut(x,c(4,10)))
table(smart_cut(x,c(4,10),labels = ~mean(.x)))   # mean of values by interval
table(smart_cut(x,c(4,10),labels = ~mean(.y)))   # center of interval
table(smart_cut(x,c(4,10),labels = ~median(.x))) # median
table(smart_cut(x,c(4,10),labels = ~paste(
  sep="~",.y[1],round(mean(.x),2),.y[2]))) # a more sophisticated label
```


## `format_fun`

With `cutr` the user can provide any formating function through the argument `format_fun`, including the package function `format_metric` (more info on value formating in the dedicated section).

```{r}
table(smart_cut(x^6 + x/100,5,"g"))
table(smart_cut(x^6 + x/100,5,"g",format_fun = format, digits = 3))
table(smart_cut(x^6,5,"g",format_fun = signif))
table(smart_cut(x^6,5,"g",format_fun = smart_signif))
table(smart_cut(x^6,5,"g",format_fun = format_metric))
```

## more on `groups`

`groups` and `n_by_group` try to place cut points at relevant quantile positions,
we won't get the required number of groups if several quantiles fall on the same 
value, to remedy to this we can use an optimization function (the default one
will most likely be enough).


```{r}
table(smart_cut(x,3,"groups"))
table(smart_cut(x,list(3,"balanced"),"groups"))
```

the second element of the list can be a string
(which will be mapped to a predefined function) or
a cusom made 2 argument function that is applied on all possible bin combinations :
Bin size is the first argument and the cut points is the second. 
the combination that return the lowest value when passed to optim_fun will be
selected (or the first of them if the minimum is not unique). "balanced"
minimizes the variance between groups and will work for most purposes.

## `cutf` and `cutf2`

These are copies of `base::cut.default` and `Hmisc::cut2` with the difference
that the formatting function can be used freely. All the features are contained
in `smart_cut` but these functions allow users to keep the interface, and
defaults of the function they know and to modify existing code easily,
for example to leverage `format_metric` with minimal effort.
